Supplementary Report for the article *Functional diversity can facilitate the collapse of an undesirable ecosystem state*

### 2022-11-12 reports/supplement/old_supplement_0.9.1.Rmd
updated supplement

# Are the results driven by functional diversity?

During the review process we received the following comment: *I think the authors have not shown sufficiently clearly that their results are actually driven by functional diversity. Fig. 4 shows that on the collapse trajectory, the dominating functional group is itself dominated by the most tolerant strain. This is also true for the suppressed group on its recovery trajectory shortly before the tipping point. This made me wonder whether the same (or at least a very similar) results regarding the change in resilience of the system states could also be obtained with just a single strain per functional group, whose tolerance to the inhibiting substance is systematically increased. This could be tested fairly easily by re-running the simulations behind Fig. 3 with just the most tolerant (and slowest growing) strain of each functional group present.* In this section we discuss this matter.

Our first realisation was that we had not adequately stated what component of functional diversity we manipulated. We revised to state *Functional diversity has multiple components, and we manipulate only one of these, namely the range of trait space represented, which has been termed functional richness, and measured as the convex hull volume ([Villéger et al. 2008](https://esajournals.onlinelibrary.wiley.com/doi/abs/10.1890/07-1206.1)).*

Since we manipulated the range of trait space represented (sensu [Villéger et al 2008](https://esajournals.onlinelibrary.wiley.com/doi/abs/10.1890/07-1206.1)), and manipulated nothing else, we manipulated only functional diversity. Logically then, any effects of this manipulation must be due to the functional diversity manipulation. That is, regardless of the process / mechanism by which functional diversity causes an effect, it is appropriate to state that functional diversity drives the results.

The reviewer’s comment motivated us to consider two questions:

1. By what mechanism does functional diversity affect ecosystem stability in this study? 
2. Could the observed effects of functional diversity also be created by manipulations of trait values that do not alter functional diversity?

Question 1. In research concerning biodiversity and ecosystem functioning there was considerable debate about the phenomenon that high yielding species are more likely to be present in more diverse communities. This phenomenon was sometimes responsible for a large proportion of observed effects of biodiversity on ecosystem function. The debate was to a large extent resolved by recognising that biodiversity effects can result from two sources: a selection/sampling effect, and a complementarity effect. Both are now considered valid and important mechanisms by which diversity (including functional diversity) affects ecosystem functioning. We view the presence of more extreme trait values in more functionally diverse communities as an analogue of the sampling/selection effect of biodiversity, and a valid mechanism by which increased functional diversity can affect ecosystem stability. For example, let us assume for the moment that, indeed, only the single maximum tolerance strain present in each functional group of an ecosystem determines the resilience of that ecosystem. Then we would say that variation in resilience with diversity operates only by the "sampling effect". If we also assume that higher tolerance causes higher ecosystem resilience, then a more functionally diverse ecosystem will have higher maximum tolerance trait, and so higher resilience. And an ecosystem with more strains per functional group will have higher resilience (Figure \@ref(fig:samp-effect1)).

Hence, even if all of the observed effects of our functional diversity manipulation were only due to presence of one strain (e.g. the most tolerant one), this can be considered an effect of functional diversity.

@@

**Question 2. It is possible that all of the observed effects of functional diversity operate via this sampling effect, and thus could also be created when there is only one maximum tolerance strain in each functional group?.**

To test if the diversity effects on resilience were driven entirely by the most tolerant strains, we made additional simulations where functional groups contained only the strain with greatest tolerance and lowest maximum growth rate. That is, for each of the 20 levels of trait variation, we ran simulations where each functional group had only one strain, and its trait values corresponded with those of the most tolerant strain for the given level of trait variation. We also manipulated in which of the functional groups tolerance increased. For example, in a simulation to be compared with variation in only the SB group (and no variation in the CB or PB groups), the SB group contained the strain with highest tolerance, whereas CB and PB each contained the strain with mean tolerance. We then compared the tipping point positions resulting from this tolerance manipulation with the tipping point positions resulting from the diversity manipulation presented in the main text.

@@

For most of the explored combinations and amounts of functional diversity/tolerance the tipping point positions did not differ between the diversity manipulation and the tolerance manipulation (Figure \@ref(fig:max-tol1)). However, the tipping point positions did differ between the diversity manipulation and the tolerance manipulation when diversity/tolerance in the SB group was high (i.e. in the combinations SB only, CB-SB, and CB-SB-PB). We show below that

>>>>

# Underlying mechanisms

In the main text we show that trait variation can affect resilience. In this supplementary section, we discuss the mechanism(s) underlying the observed diversity effects. We first explain the two major classes of mechanisms known to drive diversity effects, and then use additional simulations to elucidate the mechanisms operating in our study.

## Complementarity versus sampling effect

Research about the relationship between biodiversity and ecosystem functioning has identified two major mechanisms by which diversity affects ecosystem functioning: the selection/sampling effect and the complementarity effect (Loreau & Hector 2001). The sampling effect occurs when more diverse communities have higher functioning because they have a higher probability of containing the best-performing species, i.e. the diversity effect is driven by only one (or few) species (Cardinale et al. 2011). In contrast, the complementarity effect occurs when higher ecosystem functioning in more diverse communities is due to niche differences or positive interactions among species, i.e. the diversity effect is driven by many species (Cardinale et al. 2011). Both effects are considered valid and important mechanisms by which diversity affects ecosystem functioning (Cardinale et al. 2006, 2011, 2012; Fargione et al. 2007).

Similarly, the complementarity effect and the sampling effect are important mechanisms by which diversity affects stability (Loreau & Mazancourt 2013; Loreau et al. 2021) and resilience (Steiner et al. 2006; Hughes & Stachowicz 2011). For example, diversity can stabilize aggregate ecosystem properties (e.g. community biomass) when species differ in their responses to environmental factors and therefore respond asynchronously to environmental fluctuations (temporal complementarity effect; Loreau & Mazancourt 2013). However, when considering ecosystem responses to directional environmental change or pulse disturbances rather than to environmental fluctuations, the sampling effect might be of particular importance (Steiner et al. 2006): more diverse systems might absorb a greater amount of change by having a higher chance of containing species with high environmental tolerance. In this scenario, the positive diversity effect on resilience is driven by one (or few) species. From a conservation perspective, however, one would nevertheless strive to maintain diversity and not only the most tolerant species because of the uncertainties associated with environmental change: we do not know with sufficient certainty which stressors will change in a given ecosystem and which traits will therefore be relevant in the future.

@@

## Do the observed diversity effects operate via the sampling effect?

We investigated if the diversity effects observed in our study were in line with the sampling effect, that is, driven entirely by the presence of the most tolerant strain. Specifically, we ran additional simulations where functional groups contained only the strain with greatest tolerance and lowest maximum growth rate. That is, for each of the 20 levels of trait variation, we made simulations where each functional group had only one strain, and its trait values corresponded with those of the most tolerant strain for the given level of trait variation. We also manipulated in which of the functional groups tolerance increased. For example, in a simulation to be compared with variation in only the SB group (and no variation in the CB or PB groups), the SB group contained the strain with highest tolerance, whereas CB and PB each contained the strain with mean tolerance. We then compared the tipping point positions resulting from this tolerance manipulation with the tipping point positions resulting from the diversity manipulation presented in the main text. If the observed diversity effects are caused by the sampling effect, we expect there to be no difference between the tolerance manipulation and the diversity manipulation.

@@

For most of the explored combinations and amounts of functional diversity/tolerance the tipping point positions did not differ between the diversity manipulation and the tolerance manipulation (Figure \@ref(fig:max-tol1)). However, the tipping point positions did differ between the diversity manipulation and the tolerance manipulation when diversity/tolerance in the SB group was high (i.e. in the combinations SB only, CB-SB, and CB-SB-PB). We show below that...(TO DO: Summarize results).
